---
title: "Provider catchment areas (isochrones)"
author: "Laura Hatt"
date: '2023-03-03'
output: html_document
---

In this RMD file, I will create a polygon "catchment area" (isochrone) for each daycare provider, representing the area reachable by a 15-minute walk. Then, I will count the number of families with young children that live within each catchment area, in order to calculate the Slot-to-Population Ratio (SPR) for each daycare provider. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, echo=FALSE}
library(sf)
library(tmap)
library(tidyverse)
library(dodgr)
library(geodist)

#load developer version of osrm - there's a bug in the CRAN version
dev_lib_path <-"/Users/laurahatt/Documents/R/dev-versions/"
remotes::install_github('riatelab/osrm', lib=dev_lib_path) 
library(osrm, lib.loc = dev_lib)

path = "/Users/laurahatt/Documents/GitHub/child-care-deserts/"
```

At this point, my analysis focuses on Lakeview. This is mainly due to computer processing constraints; otherwise I would study all of Chicago.

```{r create-buffer-around-lakeview, message=FALSE}
comm_areas <- st_read(paste0(path, "Inputs/comm_areas.geojson"))
comm_areas <- st_transform(comm_areas, "EPSG:3435")
lakeview <- comm_areas %>% filter(community == "LAKE VIEW")
lakeview_buff_3435 <- st_buffer(lakeview, 5280)  #one mile buffer
lakeview_buff_4326 <- st_transform(lakeview_buff_3435, "EPSG:4326") #change back to OSMdata CRS
```

```{r load-daycares-in-lakeview-buffer, warning=FALSE}
daycare_sf <- read_sf(paste0(path, "Shapefiles/daycare_locations.shp"))
daycare_sf_4326 <- st_transform(daycare_sf, "EPSG:4326")
daycare_lakeview <- st_intersection(daycare_sf_4326, lakeview_buff_4326)
daycare_lakeview_3435 <- st_transform(daycare_lakeview, "EPSG:3435")
```

```{r create-15-min-walk-isochrones-around-daycare-providers}
isochrones <- list()
for (row in seq(1,nrow(daycare_lakeview), by=1)) {
  iso <- osrmIsochrone(loc = daycare_lakeview[row,], breaks = 15, osrm.profile="foot")
  iso['PrvdrID'] <- st_set_geometry(daycare_lakeview[row,], NULL)['PrvdrID']
  iso['DyCpcty'] <- st_set_geometry(daycare_lakeview[row,], NULL)['DyCpcty']
  isochrones[[row]] <- iso
  print(row)
}

isochrones_df = do.call(rbind, isochrones)
isochrones_3435 <- st_transform(isochrones_df, "EPSG:3435")
```

```{r load-families-as-points-shapefile}
fams_as_pts <- read_sf(paste0(path, "Shapefiles/families_as_points_lakeview.shp"))
```

```{r plot-families-and-provider-isochrones}
tmap_mode("view")
tm_shape(st_buffer(lakeview, 5280*1.5)) + tm_borders() +
  tm_shape(lakeview) + tm_polygons(alpha=0.25) +
  tm_shape(fams_as_pts) + tm_dots(col="darkred", size=0.01) +
  tm_shape(isochrones_3435) + tm_polygons(alpha=0.25) +
  tm_shape(daycare_lakeview_3435) + tm_dots(col="darkgreen", size=0.25) 
```

```{r count-families-in-each-isochrone}
num_fams_in_iso <- list()
for (row in seq(1,nrow(isochrones_3435), by=1)) {
  fams_in_iso <- lengths(st_intersects(fams_as_pts, isochrones_3435[row,])) > 0 
  fams_in_iso <- fams_as_pts[fams_in_iso,]
  num_fams <- nrow(fams_in_iso)
  num_fams_in_iso <- append(num_fams_in_iso, num_fams)
}
num_fams_in_iso <- unlist(num_fams_in_iso)
isochrones_3435['num_fams'] <- num_fams_in_iso
```

Note that I assume 1.94 young children per family, per:
https://www.statista.com/statistics/718084/average-number-of-own-children-per-family/#:~:text=The%20typical%20American%20picture%20of,18%20per%20family%20in%201960.&text=If%20there's%20one%20thing%20the,is%20known%20for%2C%20it's%20diversity.

```{r estimate-slot-to-population-ratio-for-each-provider}
isochrones_3435['num_child'] <- round(st_set_geometry(isochrones_3435, NULL)['num_fams']*1.94,0)
isochrones_3435['SPR'] <- st_set_geometry(isochrones_3435, NULL)['DyCpcty'] / 
  st_set_geometry(isochrones_3435, NULL)['num_child']
isochrones_3435
```

```{r save-isochrones-as-shapefile}
write_sf(obj=isochrones_3435, dsn=paste0(path, "Shapefiles/provider_isochrones.shp"))
```

