---
title: "Families with children under 6, by census block"
author: "Laura Hatt"
date: '2023-03-05'
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(sf)
library(tidycensus)

path = "/Users/laurahatt/Documents/GitHub/child-care-deserts/"
```

```{r load-tidycensus-API-key}
#readRenviron("~/.Renviron")`
Sys.getenv("CENSUS_API_KEY") 
```

```{r select-decennial-variables}
census_2010_vars <- load_variables(2010, "sf1", cache = TRUE)

census_2010_vars_2 <- census_2010_vars %>% 
  filter(concept == "HOUSEHOLDS BY PRESENCE OF PEOPLE UNDER 18 YEARS BY HOUSEHOLD TYPE BY AGE OF PEOPLE UNDER 18 YEARS") %>%
  select(c("name", "label"))
```

After exploring the decennial census data, I found a table enumerating the number of households per census block, disaggregated by the following:

* whether or not a household has children
* whether or not a household with children is a "family"
* whether a household is led by a married male and female, a male householder, or a female householder
* whether all the children are under 6 years old, some of the children are under 6 and others are 6-17, or all the children are 6-17.

I extracted the following variables:

* P020005 - with children, family, male and female householders, under 6 only
* P020006 - with children, family, male and female householders, under 6 and 6-17
* P020010 - with children, family, male householder, under 6 only
* P020011 - with children, family, male householder, under 6 and 6-17
* P020014 - with children, family, female householder, under 6
* P020015 - with children, family, female householder, under 6 and 6-17
* P020019 - with children, non-family, male householder, under 6 only
* P020020 - with children, non-family, male householder, under 6 and 6-17
* P020023 - with children, non-family, female householder, under 6 only
* P020024 - with children, non-family, female householder, under 6 and 6-17


```{r extract-decennial-data}
families_by_block <- get_decennial(
  geography = "block", 
  state = "IL",
  county = "COOK",
  variables = c(fam_mf_6 = "P020005",   #children under 6 only
                fam_mf_17 = "P020006",  #children under 6 AND children 6-17
                fam_m_6 = "P020010", 
                fam_m_17 = "P020011", 
                fam_f_6 = "P020014", 
                fam_f_17 = "P020015",
                nfam_m_6 = "P020019",
                nfam_m_17 = "P020020",
                nfam_f_6 = "P020023",
                nfam_f_17 = "P020024"), 
  year = 2010,
  geometry = FALSE
)
```

```{r transform-decennial-data}
families_by_block_tot <- families_by_block %>%
  spread(variable, value) %>%
  mutate(tot_fam = rowSums(cbind(fam_mf_6,
                                 fam_mf_17,
                                 fam_m_6,
                                 fam_m_17,
                                 fam_f_6,
                                 fam_f_17,
                                 nfam_m_6,
                                 nfam_m_17,
                                 nfam_f_6,
                                 nfam_f_17)))
```

```{r get-block-geometry}
blockShp <- get_decennial(geography = 'block', 
                          variables = "P001001",
                          year = 2010,
                          state = "IL",
                          county = "COOK",
                          geometry = TRUE) 

blockShp1 <- blockShp %>% 
  spread(variable, value) %>%
  select(GEOID, P001001, geometry) %>%
  st_transform("EPSG:3435")
```

```{r add-geometry-to-decennial-data}
families_by_block_sf <- merge(blockShp1, families_by_block_tot, 
                              by.x = 'GEOID', by.y = 'GEOID', all.x = TRUE)
```

```{r save-data}
write_sf(obj=families_by_block_sf, dsn="Shapefiles/families_by_block.shp")
```

